---
title: "Lab 10: Distances and the Border Zone"
author: Ava Zelenz
format: 
  html:
    code-fold: true
    toc: true
execute: 
  echo: true
---



```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

library(devtools)
library(remotes)

# Data
devtools::install_github("mikejohnson51/AOI")
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

### Question 1

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

remotes::install_github("mikejohnson51/AOI")

```

#### Get USA state boundaries

```{r}
state_conus <- aoi_get(state = 'conus')
states_conus_proj <- st_transform(state_conus, crs = 5070)
st_crs(states_conus_proj)
```

#### Get country boundaries for Mexico, the United States of America, and Canada

```{r}
countries <- aoi_get(country = c("MX", "CA", "USA"))
countries_proj <- st_transform(countries, crs = 5070)
st_crs(countries_proj)
```

#### Get city locations from the CSV file

```{r}
readr::read_csv("data/simplemaps_uscities_basicv1.90/uscities.csv")
```

```{r}
(cities = readr::read_csv("data/simplemaps_uscities_basicv1.90/uscities.csv") |> 
  select(city, state_name, county_name, population, lat, lng) )
#> # A tibble: 31,254 × 6
#>    city         state_name           county_name         population   lat    lng
#>    <chr>        <chr>                <chr>                    <dbl> <dbl>  <dbl>
#>  1 New York     New York             Queens                18832416  40.7  -73.9
#>  2 Los Angeles  California           Los Angeles           11885717  34.1 -118. 
#>  3 Chicago      Illinois             Cook                   8489066  41.8  -87.7
#>  4 Miami        Florida              Miami-Dade             6113982  25.8  -80.2
#>  5 Houston      Texas                Harris                 6046392  29.8  -95.4
#>  6 Dallas       Texas                Dallas                 5843632  32.8  -96.8
#>  7 Philadelphia Pennsylvania         Philadelphia           5696588  40.0  -75.1
#>  8 Atlanta      Georgia              Fulton                 5211164  33.8  -84.4
#>  9 Washington   District of Columbia District of Columb…    5146120  38.9  -77.0
#> 10 Boston       Massachusetts        Suffolk                4355184  42.3  -71.1
#> # ℹ 31,244 more rows
```

### Question 2:

#### 2.1 - Distance to USA Border (coastline or national) (km)

```{r}
(state_borders = st_cast(states_conus_proj, "MULTILINESTRING"))
#> Geometry set for 1 feature 
#> Geometry type: MULTILINESTRING
#> Dimension:     XY
#> Bounding box:  xmin: -106.1954 ymin: 40.25778 xmax: -104.9431 ymax: 40.99844
#> Geodetic CRS:  WGS 84
#> MULTILINESTRING ((-105.6533 40.26046, -105.6094...

cities_sf <- st_as_sf(cities, coords = c("lng", "lat"), crs = 5070)

cities_sf$dist_to_border_m <- st_distance(cities_sf, state_borders) |>
  apply(1, min)

farthest_cities <- cities_sf |>
  st_drop_geometry() |>
  arrange(desc(dist_to_border_m)) |>
  select(city, state_name, dist_to_border_m) |>
  slice_head(n = 5)

# Convert meters to kilometers for readability
farthest_cities <- farthest_cities |>
  mutate(distance_km = round(as.numeric(dist_to_border_m) / 1000, 1)) |>
  select(city, state = state_name, distance_km)

library(flextable)

flextable(farthest_cities)
```

#### 2.2 - Distance to States (km)

```{r}
state_borders_preserved <- st_boundary(states_conus_proj)


cities_sf <- st_as_sf(cities, coords = c("lng", "lat"), crs = 4326)
cities_proj <- st_transform(cities_sf, crs = 5070)

all_state_borders <- do.call(c, st_geometry(state_borders_preserved))


# Combine boundaries as an sf object (correctly preserving CRS)
all_state_borders <- st_union(state_borders_preserved)  # still sf
all_state_borders <- st_sf(geometry = all_state_borders)  # ensure it's an sf object

# Confirm both are using EPSG:5070
st_crs(cities_proj) == st_crs(all_state_borders)


cities_proj$dist_to_border_m <- st_distance(cities_proj, all_state_borders) |>
  apply(1, min)


```

```{r}
farthest_cities <- cities_proj %>%
  st_drop_geometry() %>%
  mutate(distance_km = round(as.numeric(dist_to_border_m) / 1000, 1)) %>%
  arrange(desc(distance_km)) %>%
  select(city, state = state_name, distance_km) %>%
  slice_head(n = 5)

flextable(farthest_cities)

```

#### 2.3 - Distance to Mexico (km)

```{r}
mexico <- countries %>% filter(sovereignt == "Mexico")

cities_proj <- st_transform(cities_sf, crs = 5070)
mexico_proj <- st_transform(mexico, crs = 5070)

mexico_border <- st_boundary(mexico_proj)

cities_proj$dist_to_mexico_m <- st_distance(cities_proj, mexico_border) |> 
  apply(1, min)

mex_farthest_cities <- cities_proj %>%
  st_drop_geometry() %>%
  mutate(dist_to_state_km = round(as.numeric(dist_to_mexico_m) / 1000, 1)) %>%
  arrange(desc(dist_to_state_km)) %>%
  select(city, state = state_name, distance = dist_to_state_km) %>%
  slice_head(n = 5)

flextable(mex_farthest_cities)

```

#### 2.4 - Distance to Canada (km)

```{r}
canada <- countries %>% filter(name_long == "Canada")


canada_proj <- st_transform(canada, crs = 5070)

canada_border <- st_boundary(canada_proj)

cities_proj$dist_to_canada_m <- st_distance(cities_proj, canada_border) |>
  apply(1, min)

cities_proj$dist_to_state_m <- st_distance(cities_proj, state_borders) |>
  apply(1, min)

can_farthest_cities <- cities_proj %>%
  mutate(dist_to_state_km = round(as.numeric(dist_to_canada_m) / 1000, 1)) %>%
  arrange(desc(dist_to_state_km)) %>%
  select(city, state = state_name, distance = dist_to_state_km) %>%
  slice_head(n = 5)

flextable(can_farthest_cities)

```

### Question 3

#### 3.1 Data

```{r}
ggplot() +
  geom_sf(data = countries_proj, fill = "lightgray", color = "black", lty = "dotted", size = 0.3) +
  geom_sf(data = states_conus_proj, fill = NA, color = "darkblue", lty = "solid", size = 0.5) +
  geom_sf(data = cities_proj, aes(size = population), color = "red") +
  geom_label_repel(data = cities_proj %>% slice_max(population, n = 10),
                   aes(label = city, geometry = geometry),
                   stat = "sf_coordinates",
                   size = 2,
                   min.segment.length = 0,
                   box.padding = 0.4,
                   color = "black",
                   max.overlaps = Inf
                   ) +
  theme_classic() +
  labs(title = "Top 10 Most Populated Cities in the US", 
       subtitle = "With CONUS state boundaries and neighboring countries") +
  theme_classic()
```

#### 3.2 City Distance from the Border

```{r}
cities_proj$dist_to_border_m <- st_distance(cities_proj, state_borders) |> apply(1, min)

cities_proj$dist_to_border_km <- as.numeric(cities_proj$dist_to_border_m) / 1000

five_farthest_cities <- cities_proj |>
  slice_max(dist_to_border_km, n = 5)
```

```{r}

ggplot() +
  geom_sf(data = cities_proj, aes(color = dist_to_border_km), size = 1) +
  scale_color_viridis_c(name = "Distance to Border (km)") +
  geom_sf(data = five_farthest_cities, color = "red", size = .1) +
  ggrepel::geom_label_repel(
    data = five_farthest_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 2,
    color = "black",
    max.overlaps = Inf
  ) +
  theme_minimal() +
  labs(
    title = "Distance of USA Cities from National Border",
    subtitle = "Top 5 Farthest Cities Highlighted in Red",
    x = NULL, y = NULL
  )

```

#### 3.3 City Distance from Nearest State

```{r}
state_farthest_cities <- cities_proj |>
  st_drop_geometry() |>
  mutate(dist_to_border_km = round(dist_to_border_km, 1)) |>
  arrange(desc(dist_to_border_km)) |>
  select(city, state_name, dist_to_border_km) |>
  slice_head(n = 5)

farthest_cities_coords <- st_coordinates(cities_proj)[cities_proj$city %in% state_farthest_cities$city, ]


ggplot() +
  geom_sf(data = state_borders, fill = "lightblue", color = "black") + 
    geom_sf(data = cities_proj, aes(color = dist_to_border_km), size = 2) +
    scale_color_viridis_c(option = "C", name = "Distance to Border (km)") +
  geom_label_repel(
    data = state_farthest_cities,
    aes(x = farthest_cities_coords[, 1], y = farthest_cities_coords[, 2], label = city),
    size = 3,
    box.padding = 0.5,
    point.padding = 0.5,
    max.overlaps = 10,
    segment.color = 'black',
    segment.size = 0.5
  ) +
  theme_minimal() +
  ggtitle("USA Cities Colored by Distance from State Borders") +
  theme(legend.position = "right")


```

#### 3.4 Equidistance boundary from Mexico and Canada

```{r}
new_cities_proj <- cities_proj %>%
  mutate(
    dist_to_mexico_km = dist_to_mexico_m / 1000,  # Convert Mexico distance to km
    dist_to_canada_km = dist_to_canada_m / 1000   # Convert Canada distance to km
  )

another_cities_proj <- new_cities_proj |>
  mutate(abs_diff = abs(dist_to_mexico_km - dist_to_canada_km))

highlight_cities <- another_cities_proj |>
  filter(abs_diff <= 100)

most_populous_cities <- highlight_cities %>%
  arrange(desc(population)) %>%
  slice_head(n = 5)

ggplot() +
  geom_sf(data = state_borders, fill = "lightblue", color = "black") +
  geom_sf(data = cities_proj, aes(color = dist_to_border_km), size = 2) +
  scale_color_viridis_c(option = "C", name = "Distance to Border (km)") +
    geom_sf(data = highlight_cities, aes(color = dist_to_border_km), size = 3, shape = 21, fill = "red") +
  geom_label_repel(
    data = most_populous_cities,
    aes(x = st_coordinates(most_populous_cities)[city %in% most_populous_cities$city, 1], 
        y = st_coordinates(most_populous_cities)[city %in% most_populous_cities$city, 2], 
        label = city),
    size = 3,
    box.padding = 0.5,
    point.padding = 0.5,
    max.overlaps = 10,
    segment.color = 'black',
    segment.size = 0.5
  ) +
  
  theme_minimal() +
  ggtitle("USA Cities Equidistant to the Mexican and Canadian Borders") +
  theme(legend.position = "right")
```
### Question 4
#### 4.1 Quantifing Border Zone
```{r}
border_buffer <- st_buffer(state_borders, dist = 160000)

cities_in_border_zone <- st_filter(cities_proj, border_buffer)

num_cities_in_zone <- nrow(cities_in_border_zone)

pop_in_zone <- sum(cities_in_border_zone$population, na.rm = TRUE)
total_population <- sum(cities_proj$population, na.rm = TRUE)
percent_in_zone <- round(100 * pop_in_zone / total_population, 2)

border_zone_summary <- tibble(
  Metric = c(
    "Number of Cities in Border Zone (≤160 km)",
    "Population in Border Zone",
    "Total Population (All Cities)",
    "Percent of Population in Border Zone",
    "ACLU Estimate of Population in Zone"
  ),
  Value = c(
    num_cities_in_zone,
    format(pop_in_zone, big.mark = ","),
    format(total_population, big.mark = ","),
    paste0(percent_in_zone, "%"),
    "Approximately 66%"
  )
)
    
flextable(border_zone_summary) %>%
  autofit() %>%
  set_caption("Summary of U.S. Border Zone (Within 100 Miles / 160 km)")
```
#### 4.2
```{r}
border_zone_cities <- cities_proj %>%
  filter(dist_to_border_km <= 160)

top10_border_cities <- border_zone_cities %>%
  arrange(desc(population)) %>%
  slice_head(n = 10)

ggplot() +
  geom_sf(data = state_borders, fill = "white", color = "gray60") +
  geom_sf(data = cities_proj, color = "lightgray", size = 1, alpha = 0.5) +  # background cities
  geom_sf(data = border_zone_cities, aes(color = log10(population)), size = 0.5) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  gghighlight(dist_to_border_km <= 160, use_direct_label = FALSE) +
  geom_label_repel(
    data = top10_border_cities,
    aes(
      x = st_coordinates(top10_border_cities)[, 1],
      y = st_coordinates(top10_border_cities)[, 2],
      label = city
    ),
    size = 3,
    box.padding = 0.4,
    point.padding = 0.3,
    max.overlaps = 1000,
    segment.color = "black"
  ) +
  theme_minimal() +
  labs(
    title = "Cities Within 100-Mile (160 km) U.S. Border Zone",
    subtitle = "Top 10 Most Populous Cities in the Danger Zone Highlighted",
    caption = "Source: U.S. Cities & Borders Data"
  ) +
  theme(legend.position = "right")
```
#### 4.3 : Instead of labeling the 10 most populous cites, label the most populous city in each state within the Danger Zone.
```{r}
danger_zone_cities <- cities_proj %>%
  filter(dist_to_border_km <= 160)

top_city_per_state <- danger_zone_cities %>%
  group_by(state_name) %>%
  filter(population == max(population, na.rm = TRUE)) %>%
  ungroup()

ggplot() +
  geom_sf(data = state_borders, fill = "white", color = "gray50") +
  geom_sf(data = danger_zone_cities, aes(color = log10(population)), size = 2) +
scale_color_gradient(low = "orange", high = "darkred", name = "Log10 Population") +
  geom_label_repel(
    data = top_city_per_state,
    aes(
      x = st_coordinates(top_city_per_state)[, 1],
      y = st_coordinates(top_city_per_state)[, 2],
      label = city
    ),
    size = 1.88,
    box.padding = 0.40,
    point.padding = 0.25,
    max.overlaps = 200,
    segment.color = "black",
    force = 10
  ) +
  theme_minimal() +
  ggtitle("Most Populous City per State Within 100-Mile Border Zone") +
  theme(legend.position = "right")


summary(danger_zone_cities$population)

```




